---
title: "Incidence Paper Results"
author: "Lisa Bearpark"
format: docx
editor: source
execute:
  echo: false
  warning: false
  message: false
---

### Loading the correct data and packages
```{r}
library(tidyverse)
library(survival)
library(dplyr)
library(broom)
library(naniar)
library(simputation)
library(car)
library(readr)
library(stringr)
library(ggplot2)
library(splines)

# Set base directory
directory <- system("find ~ -name \"*- ENDO_CARE*\" -type d -depth 5 -maxdepth 5 2>/dev/null | grep CARE", intern = T)
subdir1 <- "/Data/RData/"
subdir2 <- "/Studier/Incidence_of_IE/Data/RData/"

# load(str_c(directory, subdir_rdata, file)) Use if/when weights are needed
load(str_c(directory, subdir1, "endocarditis_joined_total.RData"))
load(str_c(directory, subdir2, "ie_inc_df.RData"))

endocarditis_joined_total <- endocarditis_joined_total %>% filter(year_of_diagnosis >= 1997, year_of_diagnosis <= 2023)
  

endocarditis_joined_total <- endocarditis_joined_total %>%
  mutate(
    ie_type = factor(
      protesendokardit,
      levels = c(0, 1),
      labels = c("NVE", "PVE")
    ),
    surgery = factor(
      op_endocarditis,
      levels = c(0, 1),
      labels = c("No surgery", "Surgery")
    )
  )

```

### Missing data
```{r}
#| echo: false

endocarditis_joined_total %>% 
  miss_var_summary() %>% 
  filter(pct_miss > 0) %>% 
  print(n = Inf)

ie_inc_df %>% 
  miss_var_summary() %>% 
  filter(pct_miss > 0) %>% 
  print(n = Inf)
```


# Study population
## Total number of IE cases
```{r}
nrow(endocarditis_joined_total)
```

## Number of patients by sex
```{r}
endocarditis_joined_total %>% 
  mutate(sex = factor(sex, levels = c(0, 1), labels = c("Male", "Female"))) %>% 
  count(sex) %>% 
  mutate(percent = round(100*n / sum(n), 1))

endocarditis_joined_total %>% 
  mutate(sex = factor(sex, levels = c(0, 1), labels = c("Male", "Female"))) %>% 
  filter(year_of_diagnosis == "1997" | year_of_diagnosis == "2023") %>% 
  group_by(year_of_diagnosis) %>% 
  count(sex) %>% 
  mutate(percent = round(100*n / sum(n), 1))

```

## Number of patients by age group
```{r}
endocarditis_joined_total <- endocarditis_joined_total %>%
  mutate(age = floor(age_at_diagnosis),
    age_group = case_when(
      age < 60 ~ "<60",
      age >= 60 & age < 70 ~ "60-69",
      age >= 70 & age < 80 ~ "70-79",
      age >= 80 ~ "80+",
    )
  )

endocarditis_joined_total %>% 
  mutate(
    age_group = case_when(
      age < 60 ~ "<60",
      age >= 60 & age < 70 ~ "60-69",
      age >= 70 & age < 80 ~ "70-79",
      age >= 80 ~ "80+",
    )
  ) %>% 
  count(age_group) %>% 
  mutate(percent = round(100*n / sum(n), 1))

endocarditis_joined_total %>% 
  filter(year_of_diagnosis == "1997" | year_of_diagnosis == "2023") %>% 
  group_by(year_of_diagnosis) %>% 
  count(age_group) %>% 
  mutate(percent = round(100*n / sum(n), 1))

```

## Age distribution, summary statistics
```{r}
endocarditis_joined_total %>%
summarise(
n = n(),
min = min(age_at_diagnosis, na.rm = TRUE),
q1 = quantile(age_at_diagnosis, 0.25, na.rm = TRUE),
median = median(age_at_diagnosis, na.rm = TRUE),
q3 = quantile(age_at_diagnosis, 0.75, na.rm = TRUE),
max = max(age_at_diagnosis, na.rm = TRUE),
mean = mean(age_at_diagnosis, na.rm = TRUE),
sd = sd(age_at_diagnosis, na.rm = TRUE)
)
```




# Highest and lowest incidence rates
```{r}
incidence_year <- ie_inc_df %>%
  group_by(year) %>%
  summarise(
    cases = sum(cases),
    population = sum(population),
    incidence = (cases / population) * 100000,
    .groups = "drop"
  )

bind_rows(incidence_year %>% 
    slice_max(incidence, n= 1) %>% mutate(type = "Highest"),
incidence_year %>% 
    slice_min(incidence, n= 1) %>% mutate(type = "Lowest")) %>%  
  select(type, year, incidence)

```

# Overall percentage change in incidence during the study period

```{r}
first_year <- incidence_year %>%
  filter(year == min(year))

last_year <- incidence_year %>%
  filter(year == max(year))

pct_increase <- (last_year$incidence / first_year$incidence - 1) * 100

tibble(
Metric = "Percentage increase first to last year",
Value = sprintf("%.1f%%", pct_increase)
)

```

# Percentage change first to last year by age group
```{r}

ie_inc_df <- ie_inc_df %>%
  mutate(
    age_group = case_when(
      age < 60 ~ "<60",
      age >= 60 & age < 70 ~ "60-69",
      age >= 70 & age < 80 ~ "70-79",
      age >= 80 ~ "80+",
    ))

ie_inc_age <- ie_inc_df %>%
  group_by(year, age_group) %>%
  summarise(
    cases = sum(cases),
    population = sum(population),
    incidence = cases / population * 100000,
    .groups = "drop"
  ) %>%
  arrange(desc(cases))

inc_age_first_last <- ie_inc_age %>%
arrange(age_group, year) %>%
group_by(age_group) %>%
summarise(
first_year = first(year),
last_year = last(year),
incidence_first = first(incidence),
incidence_last = last(incidence),
pct_change = 100 * (incidence_last - incidence_first) / incidence_first
)

inc_age_first_last
  

```

# Percentage change first to last year and 2010 to last year, in proportion of PVE
```{r}

perc_ie_type <- endocarditis_joined_total %>%
count(year_of_diagnosis, ie_type, name = "cases") %>%
group_by(year_of_diagnosis) %>%
mutate(
percentage = 100 * cases / sum(cases)
) %>%
ungroup()

PVE_increase_2005_2024 <- perc_ie_type %>%
filter(ie_type == "PVE") %>%
summarise(
perc_first = percentage[which.min(abs(year_of_diagnosis))],
perc_05 = percentage[which.min(abs(year_of_diagnosis - 2005))],
perc_24 = percentage[which.max(year_of_diagnosis)],
change_pp = perc_24 - perc_05,
change_pp_f_l = perc_24 - perc_first
)


PVE_increase_2005_2024

```
